---
title: "Football Kick Project"
output: html_document
date: "2023-07-14"
---

# Setting up libraries
```{r, message=FALSE, warning=FALSE}
library(zoo)
library(pracma)
library(signal)
library(tidyverse)
library(gsignal)
library(readxl)
library(patchwork)
library(magrittr)
```

# EMG processing

## Setup plotting and load data
```{r}
data_mvc_bf <- read_excel("R_BFMVC.xlsx") |>
  rename(time = `X [s]`, signal = `R BICEPS FEMORIS: EMG 2 [V]`)
data_mvc_rf <- read_excel("R_RFMVC.xlsx") |>
  rename(time = `X [s]`, signal = `R RECTUS FEMORIS: EMG 1 [V]`)
data_mvt <- read_excel("EMG_MOVEMENT.xlsx") |>
   rename(time = `X [s]`,
          signal_rf = `R RECTUS FEMORIS: EMG 1 [V]`,
          signal_bf = `R BICEPS FEMORIS: EMG 2 [V]`
    )
```



## Raw EMG signals
```{r}
gg_mvc_bf_raw <- ggplot(data_mvc_bf, aes(time, signal)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG MVC Biceps Femoris Raw") +
  theme_light()

gg_mvc_rf_raw <- ggplot(data_mvc_rf, aes(time, signal)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG MVC Rectus Femoris Raw") +
  theme_light()

gg_mvt_bf_raw <- ggplot(data_mvt, aes(time, signal_bf)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG Movement Biceps Femoris Raw") +
  theme_light()

gg_mvt_rf_raw <- ggplot(data_mvt, aes(time, signal_rf)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG Movement Rectus Femoris Raw") +
  theme_light()

```

## Filter EMG data
```{r}
fs <- 2160
lowcut <- 20
highcut <- 450
order <- 4

# Design a Butterworth filter
bf <- butter(order, c(lowcut, highcut)/(fs/2), type= "pass", plane ="z")

# Apply the filter to the data
data_mvc_rf <- data_mvc_rf |> mutate(signal_filtered = filtfilt(bf, signal))
data_mvc_bf <- data_mvc_bf |> mutate(signal_filtered = filtfilt(bf, signal))
data_mvt <- data_mvt |> mutate(signal_bf_filtered = filtfilt(bf, signal_bf), 
                               signal_rf_filtered = filtfilt(bf, signal_rf
                                                            ))
```

## Plot filtered signals
```{r}
gg_mvc_bf_filt <- ggplot(data_mvc_bf, aes(time, signal_filtered)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG MVC Biceps Femoris Filtered") +
  theme_light()

gg_mvc_rf_filt <- ggplot(data_mvc_rf, aes(time, signal_filtered)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG MVC Rectus Femoris Filtered") +
  theme_light()

gg_mvt_bf_filt <- ggplot(data_mvt, aes(time, signal_bf_filtered)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG Movement Biceps Femoris Filtered") +
  theme_light()

gg_mvt_rf_filt <- ggplot(data_mvt, aes(time, signal_rf_filtered)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG Movement Rectus Femoris Filtered") +
  theme_light()

gg_mvc_bf_raw / gg_mvc_rf_raw
gg_mvt_bf_raw / gg_mvt_bf_raw
gg_mvc_bf_filt / gg_mvc_rf_filt
gg_mvt_bf_filt / gg_mvt_bf_filt
```

## Amplitude analysis
### Calculate RMS envelope
```{r}
# Compute the power of the signal
data_mvc_bf <- data_mvc_bf |> mutate(signal_power = signal_filtered^2)
data_mvc_rf <- data_mvc_rf |> mutate(signal_power = signal_filtered^2)
data_mvt <- data_mvt |> mutate(signal_bf_power = signal_bf_filtered^2,
                               signal_rf_power = signal_rf_filtered^2
                              )

# Apply a moving window average to smooth the signal


rms_env <- function(power) {
  window_size <- 0.1
  window_size_samples <- window_size*fs
  sqrt(stats::filter(power, rep(1 / window_size_samples, window_size_samples), method = "conv"))
}

data_mvc_bf <- data_mvc_bf |> mutate(signal_rms_env = rms_env(signal_power)) |> na.omit()
data_mvc_rf <- data_mvc_rf |> mutate(signal_rms_env = rms_env(signal_power)) |> na.omit()
data_mvt <- data_mvt |> mutate(signal_bf_rms_env = rms_env(signal_bf_power),
                               signal_rf_rms_env = rms_env(signal_rf_power)) |>
  na.omit()

data_mvc_bf <- data_mvc_bf |> mutate(time = time - min(time))
data_mvc_rf <- data_mvc_rf |> mutate(time = time - min(time))
data_mvt <- data_mvt |> mutate(time = time - min(time))
```


## Normalization and plot normalized RMSE signals
```{r}
# Normalize the signal by dividing by the MVC value
mvc_bf <- max(data_mvc_bf$signal_rms_env)
mvc_rf <- max(data_mvc_rf$signal_rms_env)

data_mvc_bf <- data_mvc_bf |> mutate(signal_normalized = signal_rms_env / mvc_bf)
data_mvc_rf <- data_mvc_rf |> mutate(signal_normalized = signal_rms_env / mvc_rf)
data_mvt <- data_mvt |> mutate(bf_normalized = signal_bf_rms_env / mvc_bf,
                               rf_normalized = signal_rf_rms_env / mvc_rf
                               )
```

## Plot the normalized RMS envelope
```{r}
gg_mvc_bf_rms <- ggplot(data_mvc_bf, aes(time, signal_normalized)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG MVC Biceps Femoris Normalized RMS Envelope") +
  theme_light()

gg_mvc_rf_rms <- ggplot(data_mvc_rf, aes(time, signal_normalized)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG MVC Rectus Femoris Normalized RMS Envelope") +
  theme_light()

gg_mvt_bf_rms <- ggplot(data_mvt, aes(time, bf_normalized)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG Movement Biceps Femoris Normalized RMS Envelope") +
  theme_light()

gg_mvt_rf_rms <- ggplot(data_mvt, aes(time, rf_normalized)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Amplitude (mV)", 
       title = "EMG Movement Rectus Femoris Normalized RMS Envelope") +
  theme_light()

gg_mvc_bf_rms / gg_mvc_rf_rms
gg_mvt_bf_rms / gg_mvt_rf_rms
```

## Resample EMG data to 60 Hz
```{r}
original_frequency <- 2160  
target_frequency <- 60  
resampling_ratio <- original_frequency / target_frequency

# Resample the normalized RMS signal vectors to the target frequency
rf_resampled <- downsample(as.vector(data_mvt$rf_normalized), resampling_ratio)
bf_resampled <- downsample(as.vector(data_mvt$bf_normalized), resampling_ratio)
time_resampled <- downsample(as.vector(data_mvt$time), resampling_ratio)

emg_data = tibble(rf = rf_resampled, bf = bf_resampled, time = time_resampled)
```

# IMU processing
## Load and clean up data
```{r}
f_sampling = 60
imu_tibia <- read.csv("IMU/PantoD_20230712_114610_587.csv", fileEncoding = "Latin1", sep = ",", header = TRUE, comment.char = "#")
imu_femur <- read.csv("IMU/PiernaD_20230712_114610_592.csv", fileEncoding = "Latin1", sep = ",", header = TRUE, comment.char = "#")
imu_tibia <- imu_tibia |> mutate(time = (PacketCounter - min(PacketCounter)) / f_sampling) |> na.omit()
imu_femur <- imu_femur |> mutate(time = (PacketCounter - min(PacketCounter)) / f_sampling) |> na.omit()

diff <- nrow(imu_femur) - nrow(imu_tibia)
if(diff > 0){
  imu_femur <- imu_femur |> slice(1 : (n() - diff))
} else if(diff < 0) {
  imu_tibia <- imu_tibia |> slice(1 : (n() - abs(diff)))  
}
```


## Filter data
```{r}

# Define the filter parameters
# Sampling frequency
fs <- 60
# Cutoff freq.
fc <- 5 
# Filter order
order <- 4

# Design the buttherworth filter
f_b <- butter(order, fc / (fs/2), type="low")

# Apply the filter 
imu_femur <- imu_femur |> mutate(across(Acc_X:Mag_Z, ~ filtfilt(f_b, .x)))
imu_tibia <- imu_tibia |> mutate(across(Acc_X:Mag_Z, ~ filtfilt(f_b, .x)))
```



# Calculating IMU angle from accelerometer, gyroscope, and magnetometer data
```{r}
# Initialize orientation estimate
imu_femur <- imu_femur |> mutate(
  roll = 0,
  pitch = 0,
  yaw = 0,
  acc_roll = 0,
  acc_pitch = 0
)
imu_tibia <- imu_tibia |> mutate(
  roll = 0,
  pitch = 0,
  yaw = 0,
  acc_roll = 0,
  acc_pitch = 0
)

# Initialize complementary filter constants
tibia_filter = tibble(alpha = 0.98, beta = 1 - 0.98)
femur_filter = tibble(alpha = 0.98, beta = 1 - 0.98)

# Loop over sensor readings
estimate_angles <- function(imu, filter_consts) {
  for (i in 1:(nrow(imu) - 1)) {
    # Estimate orientation using complementary filter
    imu$acc_roll[i] <-
      atan2(imu$Acc_Y[i], imu$Acc_Z[i]) * 180 / pi
    imu$acc_pitch[i] <-
      atan2(-imu$Acc_X[i],
            sqrt(imu$Acc_Y[i] ^ 2 + imu$Acc_Z[i] ^ 2)) * 180 / pi
    
    imu$roll[i + 1] <-
      filter_consts$alpha * (imu$roll[i] + imu$Gyr_X[i] * 1 / fs) + filter_consts$beta * imu$acc_roll[i]
    imu$pitch[i + 1] <-
      filter_consts$alpha * (imu$pitch[i] + imu$Gyr_Y[i] * 1 /
                              fs) + filter_consts$beta * imu$acc_roll[i]
    imu$yaw[i + 1] <-
      imu$yaw[i] + imu$Gyr_Z[i] * 1 / fs
  }
  
  imu
}

imu_tibia <- estimate_angles(imu_tibia, tibia_filter)
imu_femur <- estimate_angles(imu_femur, femur_filter)


```

## Graph angle estimates
```{r}
gg_femur_roll <- ggplot(imu_femur, aes(time, roll)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Femur roll") +
  theme_light()

gg_femur_pitch <- ggplot(imu_femur, aes(time, pitch)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Femur pitch") +
  theme_light()

gg_femur_yaw <- ggplot(imu_femur, aes(time, yaw)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Femur yaw") +
  theme_light()

gg_tibia_roll <- ggplot(imu_tibia, aes(time, roll)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Tibia roll") +
  theme_light()

gg_tibia_pitch <- ggplot(imu_tibia, aes(time, pitch)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Tibia pitch") +
  theme_light()

gg_tibia_yaw <- ggplot(imu_tibia, aes(time, yaw)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Tibia yaw") +
  theme_light()

gg_femur_roll / gg_femur_pitch / gg_femur_yaw
gg_tibia_roll / gg_tibia_pitch / gg_tibia_yaw
```


## Convert angles to rotation matrix
```{r}
anglesToRotationMatrix <- function(roll, pitch, yaw) {
  # Convert angles to radians
  roll_rad <- roll * pi / 180
  pitch_rad <- pitch * pi / 180
  yaw_rad <- yaw * pi / 180
  
  # Calculate sin and cos values
  sr <- sin(roll_rad)
  cr <- cos(roll_rad)
  sp <- sin(pitch_rad)
  cp <- cos(pitch_rad)
  sy <- sin(yaw_rad)
  cy <- cos(yaw_rad)
  
  # Calculate rotation matrix
  R <- matrix(0, nrow = 3, ncol = 3)
  R[1, 1] <- cp * cy
  R[1, 2] <- cp * sy
  R[1, 3] <- -sp
  R[2, 1] <- sr * sp * cy - cr * sy
  R[2, 2] <- sr * sp * sy + cr * cy
  R[2, 3] <- sr * cp
  R[3, 1] <- cr * sp * cy + sr * sy
  R[3, 2] <- cr * sp * sy - sr * cy
  R[3, 3] <- cr * cp
  
  return(R)
}


# Function to calculate the angle between two sensors ( funcion para calcular el angulo entre dos vectores )
calculateAngleBetweenSensors <- function(roll1, pitch1, yaw1, roll2, pitch2, yaw2) {
  n <- length(roll1)  # Number of angles
  
  # Initialize result vectors
  roll_relative <- numeric(n)
  pitch_relative <- numeric(n)
  yaw_relative <- numeric(n)
  
  for (i in 1:n) {
    # Convert roll, pitch, and yaw angles to rotation matrices
    R1 <- anglesToRotationMatrix(roll1[i], pitch1[i], yaw1[i])
    R2 <- anglesToRotationMatrix(roll2[i], pitch2[i], yaw2[i])
    
    # Calculate the relative rotation matrix
    R_relative <- solve(R2) %*% R1
    
    # Extract roll, pitch, and yaw angles from the relative rotation matrix
    roll_relative[i] <- atan2(R_relative[2, 3], R_relative[3, 3])
    pitch_relative[i] <- -asin(R_relative[1, 3])
    yaw_relative[i] <- atan2(R_relative[1, 2], R_relative[1, 1])
  }
  
  return(list(roll_relative = roll_relative, pitch_relative = pitch_relative, yaw_relative = yaw_relative))
}


# Call the function with the given input vectors # Call the function with the given input vectors (Llamar a la función con los vectores de entrada dados) 

femur_roll <- as.matrix(imu_femur$roll)
femur_pitch <- as.matrix(imu_femur$pitch)
femur_yaw <- as.matrix(imu_femur$yaw)

tibia_roll <- as.matrix(imu_tibia$roll)
tibia_pitch <- as.matrix(imu_tibia$pitch)
tibia_yaw <- as.matrix(imu_tibia$yaw)

result <- calculateAngleBetweenSensors(femur_roll,
                                       femur_pitch,
                                       femur_yaw,
                                       tibia_roll,
                                       tibia_pitch,
                                       tibia_yaw
                                      )

relative_angles = tibble(roll = result$roll_relative * 180/pi,
                         pitch = result$pitch_relative * 180/pi,
                         yaw = result$yaw_relative * 180/pi,
                         time = imu_tibia$time
                        )
```

## Filter relative angles
```{r}

# Define the filter parameter s
# Sampling frequency
fs <- 60
# Cutoff frequency
fc <- 20
# Filter order
order <- 4
  
# Design the buttherworth filter
f_b <- butter(order, fc/(fs/2), type="low")

# Apply the filter
relative_angles <- relative_angles |> mutate(across(!time, ~ filtfilt(f_b, .x)))
```

# Plot the filtered relative angles
```{r}
gg_roll <- ggplot(relative_angles, aes(time, roll)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Relative roll") +
  theme_light()

gg_pitch <- ggplot(relative_angles, aes(time, pitch)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Relative pitch") +
  theme_light()

gg_yaw <- ggplot(relative_angles, aes(time, yaw)) +
  geom_line(color = "blue") +
  labs(x = "Time", 
       y = "Angle (deg)", 
       title = "Relative yaw") +
  theme_light()

gg_roll / gg_pitch / gg_yaw
```